Generating Random Logic Programs Using Constraint Programming

### 2020-01-19 paper/one_column/paper.tex
New formatting

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}

@@

\newtheorem{constraint}{Constraint}

@@

Parameters:
\begin{itemize}
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$ (including zero, but assuming that at
  least one predicate has non-zero arity),
  \begin{itemize}
  \item maximum arity $\maxArity{} \coloneqq \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
  \begin{itemize}
  \item Each of them can be empty, but $|\constants{}| + |\variables{}| > 0$.
  \end{itemize}
\item a list of probabilities that are randomly assigned to clauses,
\item option to forbid all cycles or just negative cycles,
\item $\maxNumNodes{} \ge 1$: maximum number of nodes in the tree representation
  of a clause,
\item $\maxNumClauses{} \ge |\predicates{}|$: maximum number of clauses in a
  program,
\item maximum number of solutions,
\end{itemize}

@@

@@

\begin{definition}
  We define $\variable{numTrees} \in \{ 1, \dots, \maxNumNodes{} \}$ to count
  the number of trees in our representation of a clause using the
  $\variable{tree}(\variable{treeStructure}, \variable{numTrees})$\footnote{This
    constraint uses dominator-based filtering by Fages and Lorca
    \cite{DBLP:conf/cp/FagesL11}.} constraint.
\end{definition}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{align*}
    \variable{count}(i, \variable{treeStructure}_{-i}) = 0 &\iff \variable{treeValues}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    \variable{count}(i, \variable{treeStructure}_{-i}) = 1 &\iff \variable{treeValues}[i].\variable{name} = \neg,\\
    \variable{count}(i, \variable{treeStructure}_{-i}) > 1 &\iff \variable{treeValues}[i].\variable{name} \in \{ \land, \lor \},
  \end{align*}
  where $\variable{treeStructure}_{-i}$ denotes array $\variable{treeStructure}$
  with position $i$ skipped.
\end{constraint}
Each constraint corresponds to node $i$ having no children, one child, and
multiple children, respectively.

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{treeStructure}[i] \ne i \implies
    \variable{treeValues}[i].\variable{name} \ne \top.
  \]
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{headsOfClauses}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodiesOfClauses}[i].\variable{numNodes} = 1,
  \]
  and
  \[
    \variable{bodiesOfClauses}[i].\variable{treeValues}[0].\variable{name} =
    \top.
  \]
\end{constraint}

@@

\begin{definition}
  We define $\arrayd{introductions}{|\variables{}|}{\{ 0, \dots, N \}}$ such
  that for $v \in \variables{}$,
  \[
    \variable{introductions}[v] = \begin{cases}
      1 + \min \variable{occurrences}[v] & \text{if }
      \variable{occurrences}[v] \ne \emptyset\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

@@

\begin{constraint}
  $\variable{allDifferentExcept0}(\variable{introductions})$.
\end{constraint}

@@

\begin{definition}
  We define an auxiliary set variable
  \[
    \variable{potentialIntroductions} \subseteq \{ 0, \dots, N \}
  \]
  by: for $i = 0, \dots, \maxNumNodes{} - 1$, if
  $\variable{treeValues}[i].\variable{name} \not\in \predicates{}$, then
  \[
    \variable{potentialIntroductions} \cap \{ \maxArity{} \times (i + 1) + j + 1
    \mid j = 0, \dots, \maxArity{} - 1 \} = \emptyset.
  \]
\end{definition} % TODO: explain these numbers
In other words, if the node is not a predicate, a variable cannot be introduced
as one of its `arguments'.

\begin{constraint}
  For $v \in \variables{}$, $\variable{introductions}[v] \in
  \variable{potentialIntroductions}$.
\end{constraint}

@@

Propagation for independence:
\begin{itemize}
\item Look up the dependencies of both predicates. For each pair of
  matching dependencies:
  \begin{itemize}
  \item If both are determined, fail.
  \item If one is determined, the selected edge of the other must not
    exist.
  \end{itemize}
\end{itemize}

% TODO: Discuss that the algorithms work in case there are multiple 'paths'
% without any issues.
% TODO: make sure my definitions and examples are consistent with loops

@@

\begin{algorithm}[t]
  \KwData{a constraint model for a logic program $\mathscr{P}$ amid execution}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with all bodies and all predicates in heads fully
  determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \lIf{\hasNegativeCycles($G_{\mathscr{R}}$)}{\Return{\textsc{false}}}
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for negative cycles}
  \label{alg:negative_cycles}
\end{algorithm}
\footnotetext{Due to the definition of independence, the arguments of the head
  atom need not be determined.}

@@

\section{Empirical Performance}
% TODO: need a lot more writing

@@

>>>>

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage{siunitx}

@@

\newtheorem{constraint}{Constraint}
\let\definition\relax
\spnewtheorem{definition}{Definition}{\bfseries}{\rmfamily}

@@

Parameters:
\begin{itemize}
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$ (including zero, but assuming that at
  least one predicate has non-zero arity),
  \begin{itemize}
  \item \emph{maximum arity} $\maxArity{} \coloneqq \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
  \begin{itemize}
  \item Each of them can be empty, but $|\constants{}| + |\variables{}| > 0$.
  \end{itemize}
\item a list of probabilities that are randomly assigned to clauses,
\item option to forbid all cycles or just negative cycles,
\item $\maxNumNodes{} \ge 1$: maximum number of nodes in the tree representation
  of a clause,
\item $\maxNumClauses{} \ge |\predicates{}|$: maximum number of clauses in a
  program,
\end{itemize}

@@

A random selection of probabilities is added at the end in case a
probabilistic program is desired. Regular logic programs can be generated by
setting the list of probabilities to $\{ 1 \}$.

@@

\begin{definition}
  Let $\variable{numTrees} \in \{ 1, \dots, \maxNumNodes{} \}$ be the number of
  trees in $\variable{treeStructure}$, defined using the $\variable{tree}$
  constraint \cite{DBLP:conf/cp/FagesL11}.
\end{definition}

@@

% TODO: could be improved by only considering positions >i
\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{align*}
    \variable{count}(i, \variable{treeStructure}_{-i}) = 0 &\iff \variable{treeValues}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    \variable{count}(i, \variable{treeStructure}_{-i}) = 1 &\iff \variable{treeValues}[i].\variable{name} = \neg,\\
    \variable{count}(i, \variable{treeStructure}_{-i}) > 1 &\iff \variable{treeValues}[i].\variable{name} \in \{ \land, \lor \},
  \end{align*}
  where $\variable{treeStructure}_{-i}$ denotes array $\variable{treeStructure}$
  with position $i$ omitted.
\end{constraint}
Each of the three equivalences in \cref{constraint:node_types} corresponds to
node $i$ having no children, one child, and multiple children, respectively.

\begin{constraint}
  Only root nodes can have $\top$ as the value. For $i = 0, \dots,
  \maxNumNodes{} - 1$,
  \[
    \variable{treeStructure}[i] \ne i \implies
    \variable{treeValues}[i].\variable{name} \ne \top.
  \]
\end{constraint}

A way to `disable' a clause:
\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{headsOfClauses}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodiesOfClauses}[i].\variable{numNodes} = 1,
  \]
  and
  \[
    \variable{bodiesOfClauses}[i].\variable{treeValues}[0].\variable{name} =
    \top.
  \]
\end{constraint}

@@

We introduce an array that, for each variable, holds the position of its first
occurrence.
\begin{definition}
  Let $\arrayd{introductions}{|\variables{}|}{\{ 0, \dots, N \}}$ be such that
  for $v \in \variables{}$,
  \[
    \variable{introductions}[v] = \begin{cases}
      1 + \min \variable{occurrences}[v] & \text{if }
      \variable{occurrences}[v] \ne \emptyset\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}
Here, a value of zero means that the variable does not occur in the clause. We
want to use specifically zero for this so that we could use
\cref{constraint:diffbutzero} later. Because of this choice, the definition of
$\variable{introductions}$ shifts all indices by one.

@@

\begin{constraint} \label{constraint:diffbutzero}
  $\variable{allDifferentExcept0}(\variable{introductions})$.
\end{constraint}

@@

\begin{definition}
  We define an auxiliary set variable
  \[
    \variable{potentialIntroductions} \subseteq \{ 0, \dots, N \}
  \]
  to act as a set of possible values that $\variable{introdutions}$ can take,
  i.e., for $v \in \variables{}$, $\variable{introductions}[v] \in
  \variable{potentialIntroductions}$.
\end{definition}

\begin{constraint} \label{constraint:potentialIntroductions}
  For $i = 0, \dots, \maxNumNodes{} - 1$, if
  $\variable{treeValues}[i].\variable{name} \not\in \predicates{}$, then
  \[
    \variable{potentialIntroductions} \cap \{ \maxArity{} \times (i + 1) + j + 1
    \mid j = 0, \dots, \maxArity{} - 1 \} = \emptyset.
  \]
\end{constraint}
In simpler terms, \cref{constraint:potentialIntroductions} says that if a node
in the tree representation of a clause represents something other than a
predicate, then a variable cannot be introduced as one of its `arguments'.

@@

% TODO: make sure my definitions and examples are consistent with loops

% TODO: Discuss that the algorithms work in case there are multiple 'paths'
% without any issues.

@@

\begin{algorithm}[t]
  \KwData{a constraint model for a logic program $\mathscr{P}$ amid execution}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with its $\variable{treeStructure}$ and predicates in both body
  and head fully determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \lIf{\hasNegativeCycles($G_{\mathscr{R}}$)}{\Return{\textsc{false}}}
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for negative cycles}
  \label{alg:negative_cycles}
\end{algorithm}
\footnotetext{The arguments (whether variables or constants) are irrelevant to
  our definition of independence.}

@@

\section{Empirical Performance}
% TODO: describe the experimental setup (variables), hardware/software (CPU,
% anything else?)

@@

% TODO: consider replacing \maxNumClauses{} with max_num_additional_clauses
\begin{itemize}
\item $|\predicates{}|, |\variables{}|, |\constants{}|, \maxNumNodes{},
  \text{the number of additional clauses} \in \{ 1, 2, 4, 8, 16 \}$.
\item $\maxArity{} \in \{ 1, 2, 3, 4, 5 \}$.
\item all possible numbers of independent pairs: $0, \dots,
  \binom{|\predicates{}|}{2}$.
\end{itemize}
Repeat 10 times:
\begin{itemize}
\item Pick a random combination of $|\predicates{}|$ arities such that
  $\maxArity{}$ is occurs at least once (in an ordered way, since order doesn't
  matter).
\item Pick the required number of independent pairs (without repetition).
\item Run the solver with a \SI{60}{\second} timeout.
\end{itemize}

Results:
\begin{itemize}
\item ??\% of total running times is less than 1 s
\item ??\% timed out. All of them had $\maxNumNodes{} = 16$.
\item Out of the ten almost-identical runs with each set of parameters, the
  number of runs that timed out was always less than ??.
\item Running time (or number of nodes) as a function of various parameters.
\end{itemize}
